Localization properties, observable character and computational reliability of the linear response in open systems

### 2018-06-12 AnalyticStructure_main.tex
continuing

% In this section we review the fundamental equations that govern the linear response regime of molecular-like systems in vacuum. Due to their intrinsic nature, this class of systems is 
% characterized by a potential energy localized in a finite region around system's atoms and free boundary conditions proper of the open systems are adopted, so the wave function is in 
% principle free to explore the whole space.

Linear response can be effectively applied to the analysis of the electronic excitations in open systems. In this framework one consider the unperturbed Hamiltonian $\hnot$ and assume that 
the ground-state density $\dmnot$ has been provided. The linear response formalism allows us to evaluate the modification of the expectation value of a generic observable induced by a 
time-dependent perturbing field acting on the system. This quantity, written in the frequency domain, can be expressed through the evaluation of the \emph{linear response functional}
\be\lb{LinearResponseFunctDef1}
\braket{\delta\op O}(\omega) = \trace{\dm'(\omega)\op O} \;,
\ee
where $\op O$ represents the observable\footnote{For a sake of concreteness, we are limiting our considerations to observables that do not explicitly depend on time} under inspection 
and we have introduced the \emph{response density operator} $\dm'(\omega)$ that codifies the modification of the ground-state density induced by the perturbation.

The main objective of this section is to investigate the structure of the building blocks needed to express the response density operator. In particular, we are interested in clarifying 
the behavior of the main actors of the LR in the long range, \emph{i.e.} outside from the support domain of the potential.

@@

\subsection{Fluctuation states. A localization argument for the response density}
\label{FluctuationState}

The response density operator is expressed as the first-order variation of ground state density $\sum_{\{p\}} \ket{\psi_p} \bra{\psi_p}$, where the set $\{\ket{\psi_p}\}$ denote the 
occupied states, \emph{i.e} the subset of the bound eigenstates of $\hnot$ with energy $\eps_p$ lower than the Fermi level. Using this notation we denote the energy of the HOMO level 
as $\eps_h$ and identify the value of the first ionization potential with $|\eps_h|$. In what follows we will assume that the set of states $\{\ket{\psi_p}\}$ are described by real 
functions, when projected in the $\r$-representation.

In the contest of the LR the response density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008})
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;,
\ee
where $\op\Phi(\omega)$ represents the perturbing field. The inspection of the linear order time evolution of the eigenstates of $\hnot$ evidence that the response density acts as a transition 
operator, linking the occupied and empty subspaces of $\hnot$, and satisfies the \emph{transverse} condition  
\be\lb{RhopTransverseDef1}
\dm' = \dm_{\perp} \doteq \dmnot\dm'\op Q_0 + \op Q_0\dm'\dmnot \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in the subspace orthogonal to the occupied states of $\hnot$. Due to this fact it is possible to introduce an explicit representation of the 
response density as follows 
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p(-\omega)} + \ketbra{f_p(\omega)}{\psi_p}\right) \;.
\ee
Here we have defined the set of $\omega$-dependent \emph{fluctuation states} $\{\ket{f_p(\omega)}\}$, that span the empty subspace of $\hnot$. 
An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states are 
written as a modified Sternheimer equations \cite{mahan1980}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'])\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains, apart from the perturbing field, a further term $\op V'[\dm']$ due to the density-dependence of $\hnot$.

@@

% The interesting aspect of these equations in the perspective of the present analysis is that they allow us to present a localization argument for the fluctuation states using a procedure  
% analogous to the one depicted in section \ref{SEopenSystem}. So we split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term and express the solution of 
% \eqref{fluctuationStateEqMotion1} in the form of a self-consistent inhomogeneous Helmholtz equation
The solution of equation \eqref{fluctuationStateEqMotion1} can be casted in the form of a self-consistent inhomogeneous Helmholtz equation
\be\lb{rhoPrimeFluctuationStateDef2}
\ket{f_p(\omega)} = \op G_H(\omega+\epsilon_p)\ket{s(\omega)} \;,
\ee
where in this case the source term contains further addend apart from the unperturbed potential and reads:
\be
\ket{s(\omega)} =  \op V\ket{f_p(\omega)} + \op Q_0(\op V'(\omega) + \op \Phi(\omega))\ket{\psi_p} \;.\nn
\ee
The localization character of fluctuation states can be assessed by following the same argument discussed in section \ref{SEopenSystem} for the eigenstates of the SE. In this case 
the source term depends also from the perturbation added to the system so this analysis applies only to the class of perturbing field for which $\ket{\op\Phi\psi_p}$ is localized, when projected
in the $\r$-representation. 
It emerges that, for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level that govern the localization properties of the corresponding 
fluctuation state. For $\omega > |\eps_p|$,  the Helmholtz kernel behaves as a spherical wave and  gives raise to a $f_p(\omega,\r)$ delocalized in the whole accessible space,
whereas for $\omega < |\eps_p|$ the kernel contains an exponential damping factor and $f_p(\omega,\r)$ exhibit a bound-state behavior in the long range.

>>>>

\subsection{Fluctuation states. A localization argument for the response density}
It is evident that the arguments presented in the above section have a limited interest for the study
of Ground State quantities of molecules. By definition, all these quantities can be expressed as a 
functional which \emph{entirely} depends on bound states. As an illustration it is sufficient to recall
the Hohenberg-Kohn theorem that states that the Ground-State (GS) energy is a functional of the GS
charge density $\dmnot$. For molecular systems that are electronically stable (i.e. no anionic or metastable configurations),
such quantity is localized, in the sense of the above section. Stated otherwise, 
\emph{there is no need to express efficiently any portion of the continuum spectrum} to have a reliable GS treatment of molecules.

Let us now inspect the case of LR calculations.
%Linear response can be effectively applied to the analysis of the electronic excitations in open systems.
In this framework one consider the unperturbed Hamiltonian $\hnot$ and assume that 
the GS density $\dmnot$ is accessible and expressed in a given computational treatment. 
The linear response formalism allows us to evaluate the modification of the expectation value of a generic observable induced by a 
time (or frequency)-dependent perturbing field $\op\Phi(\omega)$ acting on the system. 
This quantity, written in the frequency domain, can be expressed through the evaluation of the \emph{linear response functional}
\be\lb{LinearResponseFunctDef1}
\braket{\delta\op O}_\Phi(\omega) = \trace{\dm'_\Phi(\omega)\op O} \;,
\ee
where $\op O$ represents the observable\footnote{For a sake of concreteness, we are limiting our considerations to observables that do not explicitly depend on time} under inspection 
and we have introduced the \emph{response density operator} $\dm'_\Phi(\omega)$ that codifies the modification of $\dmnot$ induced by the perturbation $\op\Phi(\omega)$.

%The main objective of this section is to investigate the structure of the building blocks needed to express the response density operator. In particular, we are interested in clarifying 
%the behavior of the main actors of the LR in the long range, \emph{i.e.} outside from the support domain of the potential.

@@

\label{FluctuationState}

The response density operator is expressed as the first-order variation of $\dmnot=\sum_{\{p\}} \ket{\psi_p} \bra{\psi_p}$, where the set $\{\ket{\psi_p}\}$ denote the 
occupied states, \emph{i.e} the subset of the bound eigenstates\footnote{In what follows we will assume that the set of states $\{\ket{\psi_p}\}$ are described by real 
functions, when projected in the $\r$-representation.} of $\hnot$ with energy $\eps_p$ lower than the Fermi level. Using this notation we denote the energy of the HOMO level 
as $\eps_h$ and identify the value of the first ionization potential with $|\eps_h|$.

In the contest of the LR the response density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008})
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;,
\ee
where $\op\Phi(\omega)$ represents the perturbing field. 
The inspection of the linear order time evolution of the eigenstates of $\hnot$ evidence that the response density acts as a transition 
operator, linking the occupied and empty subspaces of $\hnot$, and satisfies the \emph{transverse} condition  
\be\lb{RhopTransverseDef1}
\dm' = \dm_{\perp} \doteq \dmnot\dm'\op Q_0 + \op Q_0\dm'\dmnot \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in the subspace orthogonal to the occupied states of $\hnot$. 
Due to this fact we can introduce an explicit representation of the 
response density parametrized as
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'_\Phi(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p^\Phi(-\omega)} + \ketbra{f_p^\Phi(\omega)}{\psi_p}\right) \;.
\ee
Here we have introduced the set of $\omega$-dependent \emph{fluctuation states} $\{\ket{f_p^\Phi(\omega)}\}$, 
that have values in the \emph{unoccupied} subspace of $\hnot$. 
An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states are 
written as a modified Sternheimer equations \cite{mahan1980}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p^\Phi(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'])\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains, apart from the perturbing field, a further term $\op V'[\dm']$ due to the density-dependence of $\hnot$.

@@

% The interesting aspect of these equations in the perspective of the present analysis is that they allow us to present a localization argument for the fluctuation states using a procedure  
% analogous to the one depicted in section \ref{SEopenSystem}. So we split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term and express the solution of 
% \eqref{fluctuationStateEqMotion1} in the form of a self-consistent inhomogeneous Helmholtz equation
It is easy to see that the solution of equation \eqref{fluctuationStateEqMotion1} can be casted in the form of
a set of scattering problems:
%a self-consistent inhomogeneous Helmholtz equation
\be\lb{rhoPrimeFluctuationStateDef2}
\ket{f_p^\Phi(\omega)} = \op G_H(\omega+\epsilon_p)\ket{s_p^\Phi(\omega)} \;,
\ee
with:%here in this case the source term contains further addend apart from the unperturbed potential and reads:
\be
\ket{s_p^\Phi(\omega)} =  \op V\ket{f_p^\Phi(\omega)} + \op Q_0(\op V'(\omega) + \op \Phi(\omega))\ket{\psi_p} \;.\nn
\ee
As a stringent hypothesis, 
let us now assume that the operator $\op Q_0(\op V'(\omega) + \op \Phi(\omega))$ is, line $\op V$ nonzero on a bounded real-space domain.
%The localization character of fluctuation states can be assessed by following the same argument discussed in section \ref{SEopenSystem} for the eigenstates of the SE. In this case 
%the source term depends also from the perturbation added to the system so this analysis applies only to the class 
%of perturbing field for which $\ket{\op\Phi\psi_p}$ is localized, when projected
%in the $\r$-representation. 
It emerges that, for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level that govern the localization properties of the corresponding 
fluctuation state. For $\omega > |\eps_p|$,  the Helmholtz kernel behaves as a spherical wave and  gives raise to a $f_p(\omega,\r)$ delocalized in the whole accessible space,
whereas for $\omega < |\eps_p|$ the kernel contains an exponential damping factor and $f_p(\omega,\r)$ exhibit a bound-state behavior in the long range.
